---
title: "Intergenerational transfers of infant mortality in 19th century northern Sweden"
author: "Göran Broström, Sören Edvinsson, and Elisabeth Engberg"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
biblio-style: apalike
documentclass: article
bibliography:
- cluster.bib
- surv.bib
abstract: This study is part of an international comparative initiative where the
  same scientific question is investigated on data sets from different regions, all
  available in IDS format. It is shown that the IDS data structure indeed can deliver
  qualified data for demographic analysis (**REALLY?**). Moreover, we show that the
  propensity of infant death for a mother is inherited from mother to daughter, which
  was the scientific question to investigate.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA, fig.height = 2.8, cache = TRUE)
options(show.signif.stars = FALSE)
library(knitr)
library(eha)
library(xtable)
```

# Introduction

This article has a threefold aim, including two methodological objectives
and one demographic query. The first methodological aim is to assess the
power of the IDS (Intermediate Data Structure) database on data from DDB
(Demographic Data Base, Umeå University) for demographic analysis. An evaluation of the
data in applied research will be useful for further development of the IDS
structure and for its future use in comparative research. The second
methodological objective is to evaluate and compare different models for
statistical analysis of inter-generational transfers. The
third and main purpose is to apply the IDS structure and the statistical modelling
to a demographic study of the possible inter-generational transmission of
death in infancy.  



# The Skellefteå region

## Geography and population

The area under study is the 19th century Skellefteå region in the province
of 
Västerbotten in the northern part of Sweden. At the outset of the study it
consisted of one large rural parish, Skellefteå. By 1900 three new parishes
had been detached into separate units, but their populations are still
included in the study until 1900.  The region was large, both in terms of
area and of population. With an area of about 1700 square miles,
Skellefteå was considerably larger than most rural parishes in Sweden. It
was considered a one-day journey to travel from the northern to the
southern border, and a ride from the coast to the more remote and sparsely
populated parts of the parish in the west could take even longer,
especially in wintertime. The main part of the population was, however,
concentrated in the coastal area and in river valleys.  In the early 19th
century the population size was around 6900, and it increased rapidly during the
first half of the century. By 1850 it had reached to about 17\,000 and at the
turn of the century it had further doubled. Despite the large increase in
population, which was mainly the result of a high natural growth, the
population density on the whole remained low [@as94].

## Economy

Skellefteå was during the studied period a rural area with a mixed economy,
based on animal husbandry, forestry and sidelines such as tar and saltpeter
production. By the mid-19th century export of tar and lumber became an
increasingly important part of the economy. The majority of the farmers in
the region were smallholders and there were no large estates. Some small
sawmills were established early in the century, but before 1900,
industrialization had little impact on the local economy. In 1835,
approximately 85 percent of the population made their living from
farming. Although the distribution of economic resources was more uniform
than in several other Swedish regions, the social stratification became
more pronounced throughout the 19th century. The increasing
proletarianization was mainly a consequence of rapid population growth. The
number of farming households remained fairly stable, while the number of
landless households increased. The socio-economic development was also
influenced by two devastating subsistence crises in the region, in the
1830s and in the 1860s [@eng05].  

## Fertility and mortality

Infant mortality was comparatively low during the 19th century [@ebrb05; @edv04]. 
Fertility was high, not only by Swedish
standards, but also in European comparison and there are no
indications of family planning. Total fertility fluctuated around five
children per woman and, although fertility did decline during the
nineteenth century, the actual fertility transition occurred late in the
district [@as94; @coa86]. The rate of illegitimacy was low in comparison
with many other parts of Northern Sweden, where frequent pre-nuptial
conceptions and illegitimate births were common [@bredro02]. The illegitimacy rate
fluctuated between three and six per cent during the nineteenth century
[@as94]. 


# Data sources


The data in this study are based on information found in the Swedish church book
records, that is, birth, death, migration, and catechetical registers
[@ndj93]. 

## The catechetical registers

In the catechetical registers the clergy kept 
continuous records of all demographic events for all individuals residing in
a parish. These detailed records make it possible to follow individuals over time and to
identify their relatives and kin over time and throughout the life course. Furthermore, a rich variety of additional
information, for example on occupation, was recorded.  From the 1970s, these records have been digitized by the Demographic Data Base at Umeå University, and individual level data from more than one hundred parishes are at present available for research in the databases [POPUM and POPLINK](http://www.cedar.umu.se/). 

## POPUM and POPLINK

The POPUM and POPLINK databases differ in two respects: The time dimension and restrictions of data access. The registration in POPUM as well as POPLINK of the Skellefteå data starts in 1699 with birth records. Catechetical registers are available from 1720. The end year of the Popum database is 1900. During recent years, the POPLINK database has been created with an extension of the time span to the 1950s for the Skellefteå and Umeå regions and a generational depth of up to 12 generations [@wee16]. The information in POPLINK is however anonymized due to legal requirements, while this is not the case for POPUM. The version of the IDS utilized in this study is based on the POPUM database, thus ending at 1900.

## Restrictions

There are flaws in the 18th and early 19th century data about infant deaths: All death registers are missing before 1815, thus researchers have to rely on death information in the catechetical registers for the earlier time. Children that died young are however under-reported in the 18th century catechetical registers. Hence the analyses in this paper are restricted to the 19th century. It includes the reproductive history of women born 1826--1854 and the information on the reproductive history of their mothers. There is a possible under-estimation of infant deaths in the grandmother generation, something we comment on and analyse in the result section. The rather short time window used here have  some methodological advantages: Women are living in comparatively similar contexts, and the risk of finding some women appearing as grand-mothers, mothers as well as children is small.  


# Implementation of IDS on POPUM

The implementation of the IDS format is based on the relational database POPUM. POPUM consists of a large amount of variables, whereof some represent the literal information from the sources, while others present updated information. The structure of POPUM differs from the one of IDS, so implementing IDS consists basically of restructuring the data. DDB was one of the pioneers in the work of testing and implementing IDS, the first version was created in 2009. Thereafter, new and improved versions have been developed. In this study we have used version 3 of IDS, which is based on only a sub-section of available parishes in POPUM - Skellefteå parish from the early 18th century until 1900. Quite a lot of variables have already been constructed in POPUM and much of the problems with inconsistent and fuzzy data, as well as duplicate information, have been taken care of. These variables are constructed from a set of well-defined and documented rules. Furthermore, record linkage as well as relational linkages have already been performed, and tests have proven the linkage process to be of high quality [@webl14]. The transformation of the data from POPUM to IDS has been quite straight-forward. This does not mean that the available information is without problems, something that will be illustrated below, but it has simplified much of the transfer of data into IDS format. The sources sometimes contain missing data, which can be a problem when it comes to analysis.
In accordance with the SEDD (Scanian Economic Demographic Database) presented by Quaranta in this volume (Quaranta 2017?), the context information on linking individuals to their households causes some problems. Families and households are presented together in the catechetical registers. We find the family on the same page in the registers and an empty line separating them from the next family. Households are however difficult to identify. Servants and others are usually presented at the bottom of the page, and if there are several families/households on the same page it is not always obvious to which family they belong. Furthermore, the separation of families/households can also be difficult, particularly in urban environments where migrations were frequent making the registers more difficult to interpret. Making rules for defining households turned out to be very complicated. The information therefore primarily represent the nuclear families as households. The approach towards how household information is entered differs in some respects between SEDD and DDB.

# Statistical modeling

The main objective of this project is to apply some common models to data sets stored in the IDS format, but coming from different data bases. For this purpose, it was desired to use some complicated models to test the strength of the ID format. The scientific objective, on the other hand, is best served by much simpler statistical models. Therefore, we split the analyses into two categories, (i) those performed by all teams, and (ii) those specific to the Skellefteå data. 

## Common models

For the common part, we follow the recipe given by Quaranta (not necessarily the scientifically best) and apply *Cox regression* for estimating the effect of grandmother's infant deaths on the time from birth to death for mother's children. Thus, instead of using the direct and reasonable approach of estimating the effect of grandmother's infant deaths on the probability of surviving to age one, we are taking the detour over proportional hazards modeling of the instantaneous hazard of dying, following infants until death or the first birthdate, whatever comes first.

A step in the right direction is taken by applying a Poisson model, where the number of infant deaths of mother is regressed on grandmother's number of infant deaths.

Both these approaches suffer from a dependent-data problem: Groups of mothers in the data set have a common mother, that is, they are sisters. Another problem is that of an asymmetry between mother and grandmother: For the grandmother only, we know that at least one child survived and got own children.  

## Simple and sound models

The bad effects of the dependent-data problem may be minimized by (i) building family trees (of women), and (ii) only study *male* infant mortality.

In this approach, the first mission is to operationalize and interpret the concept of "intergenerational transfer". There are essentially two ways to think about it.

1.    There is a causal link from the IMR of a woman to her daughter. That is, if a woman experience a fatality, the the probabilty that her daughter will experience one increases (or decreases). For such a relation to be reasonable, the woman’s fatality must come before (in calendar time) her daughter experience one.

2.    There is no direct causal link as described above, but a mother and her daughter share some common properities, they are more or less prone to have fatalities or not: Either noone or both.

In statistical terms, 1. implies a regression model, while 2. would best be described by a correlation structure. We think that 2. is the most approprate way to think about it, but for the sake of testing the extraction of data via the IDS format, both implied models will be considered.

When we compare a woman's experience of infant death to her daughter's we must take into consideration the fact that by design we *know* that the woman has at least one surviving child, the daughter we are comparing her to. This induces an asymmetry that must be resolved. We do it by considering only male infant deaths. 

A second issue is to avoid dependencies in the data. Each woman, whether mother or daughter, must be present in the data set at most once. And since data are inherently paired (grandmother and mother), siblings among daughters are not allowed.  

# Data

We use the data set that is created from the IDS database with a standard extraction script [@lq16].

## The common approach

The abovementioned data set was used asis, with the exception that mothers were limited to have been born in the time interval 1826--1854.


```{r loaddata}
if (!file.exists("ef.rda")){
    con <- url("http://capa.ddb.umu.se/ef.rda")
    load(con)
    save(ef, file = "ef.rda")
    close(con)
}else{
    load("ef.rda")
}
ef$event <- as.logical(ef$event)
levels(ef$G0_InfD_cat) <- c("0", "1", "2+", "2+")
ef$mBirthdate <- as.numeric(ef$Mother_birthdate) / 365.2425 + 130 # 1840 == 0!
ef$sex <- ef$Child_sex
```

### Variables

In the Poisson approach, the response variable is Mother's number of infant 
deaths, **mDeaths**, 
and the key explanatory variable is Grandmother's number of infant deaths, **gmDeaths**. 

In the Cox regression approach, the response variable is **Age at death** instead, 
but the key explanatory variable is still **gmDeaths**.

Also included are a bunch of possible confounders, notably

*  **gmBirths** The number of grandmother's births. Is always (by design) two or larger. In the analyses it is categorized: "2", "3", "4-6", "7+".

*  **mBirthdate** Mother's birthdate expressed as years between 1 January 1 and her day of birth, minus one.

*  **mAge** Mother's age at the birth of a child. Categorized into the intervals "17-19", "20-22", ..., "41-43", "44-49".

* **parity** The birth order of a mother's child.

*  **childBirthdate** The birthdate of a mother's child.

*  **sex** The sex of a mother's child.

These variables are not explicitly reported in the analyses, bt they are always 
present as confounders in the common analyses.

```{r redefine}
ef$gmBirths <- cut(ef$G0_Births, c(1, 2.5, 3.5, 6.5, 25), labels = c("2", "3", "4-6", "7+"))
ef$gmDeaths <- ef$G0_InfD_cat
ef$gmIMR <- ef$G0_InfD / ef$G0_Births * 100
ef$mAge <- ef$enter - 30
ef$parity <- cut(ef$Birth_order, c(0.5, 1.5, 2.5, 3.5, 6.5, 25), labels = c("1", "2", "3", "4-6", "7+"))
## ef$mBirthdate : See above!
```

## The simple and sound model

Our data extraction is not well suited for a strict analysis with a minimum of 
dependency structures. What we need is *family trees* (or rather trees of *women*), 
and for each node in the tree we want the number of male births and male infant 
deaths. We must first find the "atom mothers", that is, women without mothers 
(in our data set). Since each "MotherID" also has a "GrandmotherID", we find 
the "atoms" as those grandmothers who are not present as mothers, or

```{r atoms, echo = TRUE}
gnom <- ef$GrandmotherID[!(ef$GrandmotherID %in% ef$MotherID)]
gnom <- unique(gnom)
length(gnom)
```

So we have `r length(gnom)` atom mothers, and now we want to find their children. 
Unfortunately, this information is not present, we only have their daughters who 
survive and have children. However, we have information on the number of male 
births and male infant deaths for each atom (*G0_M_Births* and *G0_M_InfD*), and 
we can choose the the first female birth (a "Mother") as a continuation of the tree.

So, first we build generation 0:

```{r gen0, echo = TRUE}
gm <- ef[!duplicated(ef$GrandmotherID), ]
gm <- gm[gm$GrandmotherID %in% gnom, ]
gm <- gm[order(gm$GrandmotherID), ]
gnom <- sort(gnom)
#
gen0 <- data.frame(id = gnom,
                   m_id = NA,
                   generation = 0,
                   birthdate = gm$Grandmother_birthdate,
                   m_births = gm$G0_M_Births,
                   m_deaths = gm$G0_M_InfD)
summary(gen0)
```                   

Then we search for the daughters (with births) of generation0:

```{r gen1, echo = TRUE}
gm <- ef[ef$GrandmotherID %in% gnom, ]
mo <- unique(gm$MotherID)
mo <- sort(mo)
tmp <- ef[ef$MotherID %in% mo, ]
tmp <- tmp[!duplicated(tmp$MotherID), ]
tmp <- tmp[order(tmp$MotherID), ]
gen1 <- data.frame(id = mo,
                   m_id = tmp$GrandmotherID,
                   generation = 1,
                   birthdate = tmp$Mother_birthdate,
                   m_births = 0,
                   m_deaths = 0
                   )
## Count births and deaths...
for (id in gen1$id){
    xx <- ef$event[ef$MotherID == id & ef$sex == "Male"]
    gen1$m_births[gen1$id == id] <- length(xx)
    gen1$m_deaths[gen1$id == id] <- sum(xx)
}
summary(gen1)
```

For each woman in *gen1* we put on information from her mother 
about number of male births and deaths. Note that in this approach we do not 
include any of the confounders described above. 

```{r putonmoth}
indx <- match(gen1$m_id, gen0$id)
gen1$Mm_births <- gen0$m_births[indx]
gen1$Mm_deaths <- gen0$m_deaths[indx]
```

We can now test the effect of mother's infant deaths
on the probability that her daughter will experience an infant death.


# Descriptive statistics of the IDS extraction

The yearly numbers of births and deaths for the *mothers in the data set* and, as a comparison, for the data in *Poplink*,  are shown in Figure \@ref(fig:numbirth).

```{r numbirth, fig.cap="Number of births and infant deaths and infant mortality by year.", fig.height = 6}
oldpar = par(mfrow = c(2, 2))
byear <- floor(ef$mBirthdate + ef$enter) + 1840
x <- table(byear)
years <- names(x)
x <- x[-length(x)]
years <- years[-length(years)]
plot(years, as.numeric(x), type = "l", xlab = "Birth/Death year", ylab = "Frequency", col = "blue", ylim = c(0, 1500), main = "Present data")
dyear <- with(ef[ef$event, ], floor(mBirthdate + exit)) + 1840
y <- table(dyear)
years <- names(y)
lines(years, as.numeric(y), type = "l", col = "red")
abline(h = 0)
text(1850, 130, "Deaths", col = "red")
text(1825, 450, "Births", col = "blue")
ss <- with(ef, cbind(0, exit - enter, event, mBirthdate + 1840 + enter))
x <- perstat(ss, seq(1801,1901, by = 1), age = c(0, 1))
plot(1801:1900, x$intensity, xlab = "year", ylab = "Infant mortality", main = "Present data", type = "l", 
     ylim = c(0, 0.4), col = "blue")
abline(h = 0)
aver <- perstat(ss, seq(1801, 1901, by = 100), age = c(0, 1))$intensity
abline(h = aver, col = "blue", lty = 2)
## POPLINK:
library(skum)
per1 <- persons[persons$frsbobtyp == 2 & persons$region == "ske", c("foddat", "doddat")]
per1$foddat <- as.numeric(tillTid(per1$foddat))
per1$doddat <- as.numeric(tillTid(per1$doddat))
per1 <- per1[!is.na(per1$foddat), ]
per1$event <- with(per1, !is.na(doddat) & doddat - foddat <= 1)
byear <- floor(per1$foddat)
x <- table(byear)
years <- as.numeric(names(x))
x <- x[years < 1903]
years <- years[years < 1903]
plot(years, as.numeric(x), type = "l", col = "blue", xlim = c(1742, 1902), ylab = "",  xlab = "Birth/Death year", ylim = c(0, 1500), main = "POPLINK")
dyear <- floor(per1$doddat[per1$event])
y <- table(dyear)
years <- as.numeric(names(y))
y <- y[years < 1903]
years <- years[years < 1903]

lines(years, as.numeric(y), type = "l", col = "red")
abline(h = 0)
text(1850, 200, "Deaths", col = "red")
text(1825, 700, "Births", col = "blue")
ss <- with(obs, cbind(enter, exit, sluttyp == 2, birthdate))
x <- perstat(ss, seq(1801, 1901, by = 1), age = c(0, 1))
aver <- perstat(ss, seq(1801, 1901, by = 100), age = c(0, 1))$intensity
plot(1801:1900, x$intensity, xlab = "year", ylab = "Infant mortality", main = "POPLINK", type = "l", 
     ylim = c(0, 0.4), col = "blue")
abline(h = 0)
abline(h = aver, lty = 2, col = "blue")
##plot(years, as.numeric(x) / as.numeric(y), xlab = "year", ylab = "Infant mortality", main = "Present data, IMR", type = "l")
par(oldpar)
```

The difference between the two data sets is of course explained by the fact that in the present data file there are restrictions on which births to include: Mother and grandmother present, grandmother must have at least two children, etc.


```{r loaddata2}
ef <- ef[ef$mBirthdate + 1840 > 1826 & ef$mBirthdate + 1840 < 1851, ]
##ef <- ef[ef$mBirthdate + 1825 > 1800, ] ## NOTE !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

##kable(summary(ef[, c("Child_sex", "Mother_birthdate", "mBirthdate", "enter", "exit", "event")]), type = "pandoc")
```

<!---
The data frame is then aggregated to one record per mother for the Poisson regression approach, where  the influence of grandmother's infant deaths on the number of her daughter's' infant deaths is studied.
--->


```{r aggrfile}
indx <- tapply(ef$MotherID, ef$MotherID)
ef$mBirths <- with(ef, tapply(MotherID, MotherID, length)[indx])
ef$mDeaths <- with(ef, tapply(event, MotherID, sum)[indx])
ef$Mexpon <- with(ef, tapply(exit - enter, MotherID, sum)[indx])
ef1 <- ef[!duplicated(ef$MotherID), ]
##kable(summary(ef1[, c("M_Births", "M_Deaths", "G0_Births", "G0_InfD_cat")]))
```

The average number of births to a mother in the data set is 
`r round(NROW(ef) / length(unique(ef$MotherID)), 2)`. The corresponding number for a grandmother is `r round(mean(ef1$G0_Births), 2)`, but keep in mind that grandmothers are restricted to have at least two births, while mothers are restricted to have at least one birth. 

## The covered time period

Our study sample consists of all mothers born 1826--1950. The distribution of their birth years and their infant mortality (by year of child death) are shown in  Figure \@ref(fig:sample). The distribution of the grandmothers' birthdates is unknown, because it is not part of the IDS retrieval. The first births ought to have appeared in the early 1800s.  



```{r sample, fig.cap = "Distribution of mother birth years and mothers' infant mortality rate by infant death year."}
oldpar <- par(mfrow = c(1, 2))
barplot(table(floor(ef1$mBirthdate + 1840)), xlab = "Year", main = "", 
col = "lightblue",ylab = "Frequency")
##abline(h = 0)

## IMR:
years <- seq(1845, 1895, by = 1)
x <- perstat(with(ef, cbind(0, exit - enter, event, mBirthdate + 1840 + enter)), years, c(0, 1))
plot(years[-1], x$intensity, type = "b", xlab = "Year", ylab = "Infant mortality", col = "blue", ylim = c(0, 0.2))
par(oldpar)
```

## Grandmothers, mothers and mother--sisters


```{r howmany}
n.m <- nrow(ef1) # No. of mothers
n.gm <- length(unique(ef1$GrandmotherID)) # No. of grandmothers
x <- table(table(ef1$GrandmotherID)) # 'Mother siblings'
```

There are `r n.m` mothers and `n.gm` grandmothers in the data, so
obviously there are many sister groups among mothers in the data. This fact
induces dependencies in the data set, which may either be a problem (using
methods assuming independence), but it may also be possible to turn this
fact into an advantage (using mixed effects models and think of
inter-generational transfer as similarity between siblings). In the latter
case the explanatory variable *gmIMR* is replaced by
clustering on grandmother.

The distribution of the sizes of sibling groups is shown in Figure \@ref(fig:sibmoth).

```{r sibmoth, fig.cap = "Distribution of sister group sizes."}
barplot(x, xlab = "Size", ylab = "Frequency")
```

```{r countgm}
m.gm <- with(ef1, sum(unique(GrandmotherID) %in% MotherID)) 
```

How many grandmothers are also mothers (and vice versa)? The answer is `r m.gm`, or 
` round(m.gm / n.gm * 100, 2)` per cent of the grandmothers. This
small amount is of no practical importance for the results.

# Results, common models
 
The first result section, with the standard models, reports the analyses that are 
the same for all the different teams. 
The data are first analyzed with Poisson regression, and thereafter with survival analysis. In the second section, with the extended models, alternative approaches and models are investigated.

In the analyses presented here we are using **R**, a free software environment for
statistical computing and graphics [@R].
The analyses and report writing were
performed in *RStudio* [@rstudio] with the aid of the **R**
package *knitr* [@knitr; @xie15].

## Poisson regression

The expected value of the number of infant deaths $D_i$ for mother No. $i, \, i = 1, \ldots, n$,
is modeled by a Poisson distribution as 
\begin{equation*}
E(D_i) = R_i e^{\boldsymbol{\beta} \mathbf{x}_i},
\end{equation*}
where $R_i$ is total risk time for mother No. $i$, $\mathbf{x}_i$ a vector of her explanatory variables, and $\boldsymbol{\beta}_i$ is the vector of regression coefficients. (For a mother with no infant deaths, the risk time is equal to her number of births.)
Formally, $R_i$ is entered into the model as an \emph{offset} after taking logs. 



```{r poisson}
options(digits = 8)
fit.p <- glm(mDeaths ~ offset(log(Mexpon)) + gmDeaths + gmBirths + mBirthdate, data = ef1, family = poisson)
fit.p0 <- glm(mDeaths ~ offset(log(Mexpon)) + gmBirths + mBirthdate, data = ef1, family = poisson)
dr <- anova(fit.p0, fit.p)
knitr::kable(dr, row.names = TRUE, digits = 2, caption = "Analysis of deviance with (1) and without (2) gmDeaths.")
```


The results are presented in two steps: First, *the statistical significance* is calculated and shown in Table \@ref(tab:poisson): The analysis-of-deviance shows a deviance of `r round(dr$Deviance[2], 2)` with two degrees of freedom, corresponding to a *p*-value of
`r round(pchisq(dr$Deviance[2], df = 2, lower.tail = FALSE), 4)`, implying a strong statistical significance.

Regarding the practical significance, see Figure \@ref(fig:poiseff).



```{r poiseff, fig.cap = "Infant mortality by grandmother's number of infant deaths (left) and grandmother's number of births(right). Comparisons made at reference value of other covariates.", include = TRUE}
cat0 <- exp(fit.p$coef[1])
cat1 <- exp(sum(fit.p$coef[1:2]))
cat2 <- exp(sum(fit.p$coef[c(1, 3)]))
oldpar <- par(mfrow = c(1, 2))
##barplot(100 * ((c(cat1, cat2) - cat0)/ cat0), names.arg = c("1 GM death", "2+ GM deaths"), 
  ##      ylim = c(0, 50), ylab = "Relative IMR (per cent)")

barplot(100 * c(cat0, cat1, cat2), names.arg = c("0", "1", "2+"), 
        ylim = c(0, 9), ylab = "IMR (per cent)", xlab = "gmDeaths")
cat0 <- exp(fit.p$coef[1])
cat1 <- exp(sum(fit.p$coef[c(1, 4)]))
cat2 <- exp(sum(fit.p$coef[c(1, 5)]))
cat3 <- exp(sum(fit.p$coef[c(1, 6)]))

##barplot(100 * ((c(cat1, cat2) - cat0)/ cat0), names.arg = c("1 GM death", "2+ GM deaths"), 
  ##      ylim = c(0, 50), ylab = "Relative IMR (per cent)")

barplot(100 * c(cat0, cat1, cat2, cat3), names.arg = c("2", "3", "4-6", "7+"), 
    ylab = "IMR (per cent)", xlab = "gmBirths")

```

The size of the effect of grandmother's infant mortality on mother's is not large, but, maybe surprisingly, the effect of her number of births is. 

## Survival analysis

The **R** package *survival* [@survival] is used, and the
explanatory and potentially confounder variables are the same as in the Poisson regression
analysis. The statistical significance of a Cox regression is shown in Table \@ref(tab:surv): The first row is the full model, the second with the variable *gmDeaths* removed, and the third with both *gmDeaths* and *gmBirths* removed. The small *p*-values show that both variables are needed in the model. 

```{r surv}
##ef$G0InfD <- ef$G0_InfD_cat
##ef$G0Births.c <- ef$G0_Births -
##ef$mAge <- ef$enter - 32
##ef$mAge2 <- (ef$mAge)^2
##ef$mAge3 <- (ef$mAge)^3
ef$mAge <- cut(ef$enter, c(seq(16.99, 43.99, by = 3), 49.99), labels = c("17-19", "20-22", "23-25", "26-28", "29-31", "32-34", "35-37", "38-40", "41-43", "44-49"))
ef$childBirthdate <- ef$mBirthdate + ef$enter
fit.c <- coxph(Surv(exit - enter, event) ~ gmDeaths + gmBirths + childBirthdate + 
                    strata(parity) + mAge, data = ef)
fit.c1 <- coxph(Surv(exit - enter, event) ~ gmBirths + childBirthdate + 
                    strata(parity) + mAge, data = ef)
fit.c2 <- coxph(Surv(exit - enter, event) ~ childBirthdate + 
                    strata(parity) + mAge, data = ef)
dr.c <- anova(fit.c, fit.c1, fit.c2)
##kable(summary(fit.c)$coef[, 1:3], caption =  "Cox regression, standard model.", booktabs = TRUE)
##ltx(fit.c, dr = dr.c, caption = "Cox regression, standard model.", 
  ##     label = Table: (\#label) tab:surv, digits = 4)
knitr::kable(dr.c, row.names = TRUE, digits = 4, caption = "Analysis of deviance: (1) full model, (2) gmDeaths removed, and (3) gmDeaths and gmBirths removed.")
```


Second, the sizes of the effects are shown in Figure \@ref(fig:effstcox). For the
covariate *gmDeaths*, the real effect on
her daughter's risk of experiencing infant deaths appears when the number
of deaths are two or larger. *gmBirths* is also
influential. 

```{r effstcox, fig.cap = "Effects of included covariates, Cox regression. The leftmost value is the reference in both panels.", fig.height = 3.3}
source("R/eff.R")
oldpar <- par(mfrow = c(1, 2))
sdp <- sqrt(diag(fit.c$var))
##gmDeaths:
##eff.fact(fit.c$coef[1:2], sdp[1:2], labs = fit.c$levels[[1]], xlab = "No. of GM deaths")
eff.fact(fit.c$coef[1:2], sdp[1:2], labs = c("0", "1", "2+"), xlab = "No. of GM deaths")
##c0 <- 0
##c1 <- 100 * (exp(fit.c$coef[1]) - 1)
##c2 <- 100 * (exp(fit.c$coef[2]) - 1)
##barplot(c(c0, c1, c2), names.arg = c("0", "1", "2+"), main = "No. of GM deaths", ylab = "Rel. Risk ##increase (per cent)", ylim = c(0, 60))
##gmBirths:
eff.fact(fit.c$coef[3:5], sdp[3:5], labs = c("2", "3", "4-6", "7+"), xlab = "No. of GM births")
##c0 <- 0
##c1 <- 100 * (exp(fit.c$coef[3]) - 1)
##c2 <- 100 * (exp(fit.c$coef[4]) - 1)
##c3 <- 100 * (exp(fit.c$coef[5]) - 1)
##barplot(c(c0, c1, c2, c3), names.arg = c("2", "3", "4-6", "7+"), main = "No. of GM births", ylab = "Rel. ##Risk increase (per cent)", ylim = c(0, 60))
## mAge:
##eff.fact(fit.c$coef[6:14], sdp[6:14], labs = fit.c$levels[[3]], xlab = ##"Mother's age")
##x <- seq(20, 45, length = 25)
##y <- exp(fit.c$coef[6] * (x - 32) + fit.c$coef[7] * (x - 32)^2) - 1
##plot(x, 100 * y, xlab = "Age", type = "h", main = "Mother's age", ylab = "Rel. Risk increase (per cent)", ##axes = FALSE, ylim = c(0, 60))
##axis(1, at = c(20, 25, 32, 40, 45))
##axis(2)
##box()
##abline(h = 0)
par(oldpar)
```

### The proportional hazards assumption

The estimated cumulative hazards functions for the strata of
*gmDeaths* are shown in Figure \@ref(fig:relcumhaz).

```{r relcumhaz, fig.cap = "Cumulative hazards by the number of grandmother's infant deaths.", fig.height = 3.2}
fit.str <- coxreg(Surv(exit - enter, event) ~ strata(gmDeaths) + gmBirths + mAge, data = ef)
plot(fit.str, col = c("black", "blue", "red"), xlab = "Age", ylab = "Cumulative hazards", lty = c(1, 2, 4))
ef_fix <- ef
save(ef_fix, file = "ef_fix.rda")
```

There is an evident deviation from the assumption of *proportional
  hazards*: However, this does not disturb the main conclusion: Two or more
grandmother infant deaths is harmful, quite in line with the results from
the Poisson regression. The reported *p*-values are cast in doubt, however.  

### Dependency structures

There are a couple circumstances that introduce dependence structures in the data: Infants being siblings share genetic and environmental unmeasurable properties, some mothers have sisters that themselves are present as mothers, thus sharing grandmother. Possible ways of handling the situation are *shared frailty models* [@aalen08} and the implementation of *robust variances* [@extcox00]. We have tried both, but neither do change the results in any noticeable way. For the shared frailty model we get a very strong frailty effect of the grouping of infants by mother (not shown, and not surprising).

<!---
```{r shared, echo = FALSE}
library(coxme)
fit.c2 <- coxph(Surv(exit - enter, event) ~ gmIMR + Child_birthyear + 
                    parity + sex, data = ef)
fit.c3 <- coxme(Surv(exit - enter, event) ~ gmIMR + Child_birthyear + parity + sex + (1 | MotherID), data = ef)
```

```{r sharedpres}
anova(fit.c2, fit.c3)
```

The estimated fixed effects coefficients are found in

```{r foundcoxef}
print(fit.c3, digits = 2)
```
--->
Our main hypothesis, there is no intergenerational transfer, is still rejected.


# Extended results

Findings that are specific for the Skellefteå project are presented here. 
First we look at the simple-and-sound model based on the family trees with only 
male infant mortality. Then, for comparison, the survival analysis is repeated 
with only male infants.  

## Simple-and-sound

The simple data can be summarized in a contingency table (Table \@ref(tab:sos)).

```{r sos}
x <- with(gen1, table(m_deaths, Mm_deaths))
col_names <- paste("gmD =", 0:5)
rownames(x) <- paste("mD = ", 0:3)
knitr::kable(x, caption = "Contingency table of mother's (mD) and grandmother's (gmD) infant deaths.",
             col.names = col_names, row.names = TRUE)
```

```{r whipr}
fit <- glm(m_deaths ~ I(Mm_deaths > 1) + I(Mm_deaths == 1), data = gen1, family = poisson)
summary(fit)
```

The conclusion is as usual: Mother's experience of at least two infant 
deaths increases daughter's probability of an infant death significantly 
(odds ratio around 1.6).

### The Poisson model

It turns out that the model fit to data is somewhat better with *number of births* as the offset (log scale). Also showing a slight improvement is to use *gmIMR* as explanatory variable rather than her *absolute number of deaths*. The *IMR* is defined as the number of infant deaths divided by the number of births. These changes also implies that the comparison *mother vs. grandmother* happens on a probability scale rather than on the intensity one.

```{r refinedpois, results = "asis"}

fit.p2 <- glm(mDeaths ~ offset(log(Mexpon)) + gmIMR + mBirthdate + parity, data = ef1, family = poisson)
dr <- drop1(fit.p2, test = "Chisq")
kable(dr, caption = "Poisson regression, grandmother's IMR. Analysis of deviance.", digits = 3, booktabs = TRUE)
```

The general conclusion is not changed: Grandmother's IMR has a strong
influence on her daughter's IMR. 

The size of the effect is shown in Figure \@ref(fig:plotimr).

```{r plotimr, fig.cap = "Probability of infant death for mother (M IMP) by IMR for grandmother (GM IMR). Poisson regression."}
x <- seq(0, 100, length = 1000)
mimr <- with(fit.p2, exp(coefficients[1] + coefficients[2] * x))
plot(x, mimr, type = "l", col = "blue", xlab = "GM_IMR (per cent)", ylab = "M_IMP", ylim = c(0, 0.20))
abline(h = 0)             
```

### Survival analysis

The same modification as in the Poisson case is introduced here.

```{r extsurv,results='asis'}
ef$childBirthdate <- ef$mBirthdate + ef$enter
fit.c2 <- coxph(Surv(exit - enter, event) ~ gmIMR + Child_birthyear + parity + sex, data = ef)
dr.c2 <- drop1(fit.c2, test = "Chisq")
##ltx(fit.c2, dr = dr.c2, caption = "Cox regression, extended model.", digits = 4, label = "tab:extsurv")
kable(dr.c2, caption = "Cox regression, extended model.", booktabs = TRUE, digits = 3)
```

The regression coefficient for *gmIMR* is `r round(fit.c2$coef[1], 3)`, highly statistically significant, as seen in Table \@ref(tab:extsurv). The other covariates are of less importance.

```{r survm}
##ef$G0InfD <- ef$G0_InfD_cat
##ef$G0Births.c <- ef$G0_Births -
##ef$mAge <- ef$enter - 32
##ef$mAge2 <- (ef$mAge)^2
##ef$mAge3 <- (ef$mAge)^3
ef$mAge <- cut(ef$enter, c(seq(16.99, 43.99, by = 3), 49.99), labels = c("17-19", "20-22", "23-25", "26-28", "29-31", "32-34", "35-37", "38-40", "41-43", "44-49"))
ef$childBirthdate <- ef$mBirthdate + ef$enter
fit.cm <- coxph(Surv(exit - enter, event) ~ I(G0_M_InfD /
                 (G0_M_Births + 0.001)) + #childBirthdate + 
                    ###strata(parity) + 
                    mAge, data = ef[ef$Child_sex == "Male", ])
dr.cm <- drop1(fit.cm, test = "Chisq")
##kable(summary(fit.c)$coef[, 1:3], caption =  "Cox regression, standard model.", booktabs = TRUE)
##ltx(fit.c, dr = dr.c, caption = "Cox regression, standard model.", 
  ##     label = Table: (\#label) tab:surv, digits = 4)
dr.cm
```

#### Other ways of eliminating sibling effects

A radical way to eliminate the sibling effect among infants is to include the firstborn for each mother. We get the results shown in Table \@ref(tab:extsurv1).

```{r extsurv1,results='asis'}
ef3 <- ef[ef$parity == 1, ]
fit.c3 <- coxph(Surv(exit - enter, event) ~ gmIMR + childBirthdate + sex, data = ef3)
dr.c3 <- drop1(fit.c3, test = "Chisq")
#ltx(fit.c3, dr = dr.c3, caption = "Cox regression, extended model with only one birth per mother.", digits = 4, label = #"tab:extsurv1")
kable(summary(fit.c3)$coef[, 1:3], caption = "Cox regression, extended model with only one birth per mother.", booktabs = TRUE, digits = 3)
```

The effect of *gmIMR* is even stronger for firstborn than generally (estimated to `r round(fit.c3$coef[1], 3)`), however, the statistical significance is weak, a logical consequence of the much lower number of infant deaths (compare Tables \@ref(tab:extsurv) and \@ref(tab:extsurv1)).


### A 2-by-2 table

In order to make it really simple, let us just record whether a mother and a grandmother experienced an infant death or not. The result is, in tabular form,

```{r twotwo}
gmD <- ef1$gmDeaths != 0
mD <- ef1$mDeaths > 0 
y <- table(mD, gmD)
zz <- fisher.test(y)
##kable(y)
y
```

<!-- $$ -->
<!-- \begin{center} -->
<!-- \begin{tabular}{l|rr|r} -->
<!--      & \multicolumn{2}{l}{\bf Mother death} & \\ \hline -->
<!-- \bf Grandmother death & No & Yes & \bf Sum \\ \hline -->
<!-- No & \Sexpr{y[1, 1]} & \Sexpr{y[1, 2]} & \Sexpr{sum(y[1, ])} \\ -->
<!-- Yes & \Sexpr{y[2, 1]} & \Sexpr{y[2, 2]} & \Sexpr{sum(y[2, ])} \\ \hline -->
<!-- \bf Sum & \Sexpr{sum(y[, 1])} & \Sexpr{sum(y[, 2])} & \Sexpr{sum(y)} \\ \hline -->
<!-- \end{tabular} -->
<!-- \end{center} -->
<!-- $$ -->

The *odds ratio* in this table is `r round(zz$estimate, 2)`, and *Fisher's exact test* [@fisher22] gives a $p$-value of `r round(zz$p.value, 4)` and a 95\% confidence interval 
(`r round(zz$conf.int, 2)`). Expressed in probabilities: If no grandmother death, then the probability of a mother death is *605 / 1576 = 0.38*, while if grandmother experienced a death the corresponding probability is *298 / 671 = 0.44*, an increase by *16%*.

An even simpler table is obtained if we do not allow siblings among mothers, that is, to each grandmother only one mother (her first-born daughter) is connected:

```{r twotwo2}
gmD <- ef1$gmDeaths[!duplicated(ef1$GrandmotherID)] != 0
mD <- ef1$mDeaths[!duplicated(ef1$GrandmotherID)] > 0 
y <- table(mD, gmD)
zz <- fisher.test(y)
#kable(y)
y
```


<!-- \begin{center} -->
<!-- \begin{tabular}{l|rr|r} -->
<!--      & \multicolumn{2}{l}{\bf Mother death} & \\ \hline -->
<!-- \bf Grandmother death & No & Yes & \bf Sum \\ \hline -->
<!-- No & \Sexpr{y[1, 1]} & \Sexpr{y[1, 2]} & \Sexpr{sum(y[1, ])} \\ -->
<!-- Yes & \Sexpr{y[2, 1]} & \Sexpr{y[2, 2]} & \Sexpr{sum(y[2, ])} \\ \hline -->
<!-- \bf Sum & \Sexpr{sum(y[, 1])} & \Sexpr{sum(y[, 2])} & \Sexpr{sum(y)} \\ \hline -->
<!-- \end{tabular} -->
<!-- \end{center} -->

Now, the *odds ratio* in this table is `r round(zz$estimate, 2)`, and *Fisher's exact test* [@fisher22] gives a *p*-value of `r round(zz$p.value, 4)` and a 95% confidence interval 
(`r round(zz$conf.int, 2)`). Expressed in probabilities: If no grandmother death, then the probability of a mother death is `r round(y[1, 2] / sum(y[1, ]), 2)`, while if grandmother experienced a death the corresponding probability is `r round(y[2, 2] / sum(y[2, ]), 2)`, an increase by 16%.

An astonishing similarity!

### The 2-by-2 table with covariates

We can of course analyze these table data by *binomial regression*, including the same covariates as earlier. Since the results from this exercise do not differ from earlier results, we refrain from showing them.

```{r getef2}
ef2 <- ef1[!duplicated(ef1$GrandmotherID), ]
ef2$gmD <- ef2$gmDeaths != "0" 
ef2$mD <- ef2$mDeaths > 0
fit <- glm(mD ~ gmD + G0_Births + mBirthdate, data = ef2, family = binomial)
#summary(fit)
```

# A sensitivity analysis

The death books for the years 1822--1830 are lost in a fire, so the only information about deaths for these years is what is found in the husförhörslängder. This means that very early deaths may not be reported. See Figure \@ref(fig:numbirth).


# Concluding remarks

Three main aims were set out for this article. The first was about
assessing the power of the IDS data format for demographic analysis, in
this case in relation to transnational comparative research.
We can confirm that applying the IDS format, and the general script made by
@lq16, has fulfilled its purposess well. The use of the script has
facilitated comparisons between countries. 
However, it is important to remember that the IDS format do not relieve the
researcher from the duty of carefully evaluating the sources and their usefulness for
the research questions raised. In the database for the Skellefteå region,
we had to restrict our studied period substantially, due to deficient data
for the 18th century, where infant deaths were severely underestimated. 
This problem can eventually be taken care of by applying clear definitions of rules on how to 
identify infant deaths, but that must be performed in a preparatory state.

Second, we have tried different analytic models, ranging from complicated
(proportional hazards models for survival times of infants)
to simple (had a woman and/or her oldest daughter an infant death or not: A
contingency table analysis). ``Complicated'' refers to ``strong
assumptions and strict model checking'': The hard part is dealing with the
complicated dependency structures arising from sisters having the same
mother and the dependency between life lengths of sibling infants. On the
other hand, ``simple'' means almost no restrictive assumptions, no
complicated dependency structures, and results
that are directly interpretable in terms of simple-to-understand probabilities.
As is often the case, simple is at least as good as complicated: The
different models give similar results, and it is easy to feel confident in
the ones with the weakest assumptions. On the other
hand, it is generally good (statistical) practice to try different approaches to the
analysis of the
same problem. You may call it *sensitivity analysis* [@co84,chapter
6] if you wish.  

Third, when it comes to the results, we find a clear association between infant
mortality among the mothers and the grandmothers.  Another question is if
this is of practical significance? The simplest model is the easiest to interpret: If grandmother has no infant deaths, then the probability for mother to have one is 38 per cent, otherwise it is 45 per cent.  


# Discussion

Our study confirms the existence of family dependency in infant mortality
risks. How can the association be explained? This relates to the old debate
of whether humans are primarily formed by nature or if nurture is
important, a division that now is becoming muddled and partly obsolete due
to recent scientific developments [@mel14]. In the
analytic approach taken here, we cannot really distinguish between purely
biological versus social and cultural factors, but there are
possible pathways. 

One obvious possible factor refers to genetics. People are differently frail and it is
reasonable that this frailty is genetically transferred. This might affect childrens' 
capacity of survival as well as factors related to the mother associated with delivery and her
the ability to take care of the newborn. There are also
other possible biological reasons. The Rh negative blood group is common in
our region, thus increasing the risk for Rh disease. It has been shown that
this condition did have some impact on perinatal health in the region, although on a
fairly restricted level. Still it can still be a component in the observed
association [@hle12]. Furthermore, a
mother and her sibling group can have experienced problematic conditions in
early childhood that may have scarred her for life, making her biologically
less fit for child birth and thereby increasing the mortality risks for her
children [@lq13].   

Can the survival of children have to do with transferred behaviours between
generations? We know that the ways children are taken care of is crucial
for their health and survival, particularly in a high mortality
regime. This can relate to local patterns of childcare, for example to the
practice of breast-feeding, but also to practices that are
transferred within families. Our immediate family is the closest social
organization for learning. In 19th century Sundsvall, mothers
moving to the region kept their childcare practices from their home region in the new
environment. Mothers from a high mortality region experienced a higher infant mortality and vice versa [@lgt04].
Another explanation for the association between generations is that they
merely reflect similar living conditions, for example the physical
environment they lived in or the belonging to the same social class. This
could be more thoroughly analyzed in models where residence and/or social
position are considered. 


Regarding the results, the main hypothesis was confirmed: The risk for a
woman to experience infant deaths is increased if her mother also had
that experience. While statistically significant, the effect size is moderate. 
We applied different models, all plausible, and got almost
identical results in all cases. 

# References 


