---
title: "Grandmother's children"
author: "Göran Broström"
date: "January 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(tidyr)
library(haven)
library(dplyr)
library(knitr)
mdy <- function(Month, Day, Year) as.Date(paste(Year, Month, Day, sep = "-"), 
                                          format = "%Y-%m-%d")
```

## Introduction

We want to find grandmother's all children, not only those that appear as 
mothers in Luciana's output. 

I will doing it by "reverse engineering", that is, I start by finding all 
grandmothers in *ef.rda* (Luciana's output converted to R format), then I look for all their kids via *INDIV_INDIV* and *INDIVIDUAL* (IDS files).

## Find the grandmothers (GM)

```{r loaddata}
load("ef_fix.rda")
ef <- ef_fix
rm(ef_fix)
GMid <- unique(ef$GrandmotherID)
Mid <- unique(ef$MotherID)
Chid <- unique(ef$Id_I) 
load("../IDS/IDS_ddb/INDIVIDUAL.rda") 
INDIVIDUAL <- INDIVIDUAL %>%
   select(Id_I, Type, Value, Day, Month, Year, Start_day, Start_month, 
           Start_year, End_day, End_month, End_year,
           Date_type, Missing)
load("../IDS/IDS_ddb/INDIV_INDIV.rda") 
INDIV_INDIV <- INDIV_INDIV %>%
    select(Id_I_1, Id_I_2, Relation) %>%
    filter(Relation == "Mother") %>% 
    select(Mother = Id_I_1, Child = Id_I_2) %>%
    filter(!duplicated(Child))
```

Now, from *INDIV_INDIV* we select all rows ('filter') with *Mother* 
corresponding to *GMid* (from *ef*).

```{r Gmchild}
GMchild <- INDIV_INDIV %>%
    filter(Mother %in% GMid)
```

These grandmothers had a lot of children:

```{r tapply}
x <- with(GMchild, tapply(Mother, Mother, length))
barplot(table(x), xlab = "No. of children to grandmother", ylab = "Frequency")
```

## Find the kids of the grandmothers

That is, the *mothers* and their siblings. From *INDIVIDUAL*, after some 
massage:

```{r get_kids_to_GM}
kidsGM <- INDIVIDUAL %>%
    filter(Id_I %in% GMchild$Child) ##%>% 
##    filter(Type %in% c("BIRTH_DATE", "DEATH_DATE", "START_OBSERVATION", 
  ##         "END_OBSERVATION"))
kidsGM$date <- with(kidsGM, mdy(Month, Day, Year))
kidsGM <- select(kidsGM, Id_I, Type, Value, date) %>%
    filter(!is.na(date))

far <- with(kidsGM, paste(Id_I, Type, Value, date, sep = ""))
kidsGM <- kidsGM[!duplicated(far), ]
                
indx <- match(kidsGM$Id_I, INDIV_INDIV$Child)
kidsGM$GMid <- INDIV_INDIV$Mother[indx]
summary(kidsGM)
```

We see that grandmother has a birth as early as 1804-01-13 and her longest 
living kid dies 1902-01-026.

We will now check the birthdate , deathdate and 'last seen' for all GMkids.
We start by sorting data by Id_I and date and then add *lopnr* and *antrec*: 

```{r rc}
source("rc.R")
kidsGM$id <- kidsGM$Id_I
kidsGM <- rc(kidsGM)
kidsGM$id <- NULL
head(kidsGM, 15)
```

Individuals with missing birthdate are removed:

```{r withbdate}
kids <- kidsGM %>%
    filter(Type == "BIRTH_DATE")
nk <- data.frame(id = kids$Id_I, GMid = kids$GMid, birthdate = kids$date)
deaths <- kidsGM %>%
    filter(Type == "DEATH_DATE")
indx <- match(nk$id, deaths$Id_I)
nk$deathdate <- deaths$date[indx]
last <- kidsGM %>%
    filter(antrec == lopnr)
indx <- match(nk$id, last$Id_I)
nk$last <- last$date[indx]
head(nk)
```

We calculate the number of days from birth to "last seen":

```{r lifetime}
td <- function(date1, date2){
    ## vectorizing...
    diff <- as.numeric(date2) - as.numeric(date1)
    diff
}

nk$exit <- with(nk, td(birthdate, last))
nk$event <- (!is.na(nk$deathdate)) & nk$exit < 366
nk$exit <- pmin(nk$exit, 365)
nk <- nk[nk$exit >=0, ] 
```

Then we introduce *period* as covariate, categorizing *birthdate* into three categories,

*birthdate < 1822-01-01*, *1822-01-01 <= birthdate < 1831-01-01*, and *1822-01-01 <= birthdate*:

```{r period}
cut1 <- as.Date("1822-01-01")
cut2 <- as.Date("1831-01-01")
nk$period <- cut(nk$birthdate, c(as.Date("1804-01-01"), cut1, cut2, as.Date("1872-12-31")))
```


## Survival analysis

We compare survival in the three time periods by plotting the cumulative hazards, 
with and without "correction".

```{r theplots, fig.height = 8}
library(eha)
nk$exit <- nk$exit + 1
fit <- coxreg(Surv(exit, event) ~ strata(period), data = nk)
par(mfrow = c(2, 1))
plot(fit, col = c("black", "red", "blue"), main = "Uncorrected", lty = 1:3, ylim = c(0, 0.3), xlab = "Days")
nk$event1 <- nk$event | nk$exit == 1
fit2 <- coxreg(Surv(exit, event1) ~ strata(period), data = nk)
plot(fit2, col = c("black", "red", "blue"), main = "Corrected", lty = 1:3, ylim = c(0, 0.3), xlab = "Days")
```


"1" is 1804-1821, "2" is 1822-1831, and "3" is 1832-1872. "Corrected" means that those with only a birthdate are regared as dead on the first day of life.

It is obvously not only a problem with under-registration very early, but up to three months of age: Mortality after that age looks "reasonable":

```{r theplots2, fig.height = 5}
##library(eha)
##nk$exit <- nk$exit + 1
nk$enter <- 0
nk2 <- age.window(nk, c(90, 366))
fit <- coxreg(Surv(enter, exit, event) ~ strata(period), data = nk2)

plot(fit, col = c("black", "red", "blue"), main = "3-12 months", lty = 1:3, ylim = c(0, 0.06), xlab = "Days")

```

A quite reasonable picture.
