---
title: "Grandmother's children"
author: "Göran Broström"
date: "January 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(tidyr)
library(haven)
library(dplyr)
library(knitr)
mdy <- function(Month, Day, Year) as.Date(paste(Year, Month, Day, sep = "-"), 
                                          format = "%Y-%m-%d")
```

## Introduction

We want to find grandmother's all children, not only those that appear as 
mothers in Luciana's output. 

I will doing it by "reverse engineering", that is, I start by finding all 
grandmothers in *ef.rda* (Luciana's output converted to R format), then I look for all their kids via *INDIV_INDIV* and *INDIVIDUAL* (IDS files).

## Find the grandmothers (GM)

```{r loaddata}
load("ef_fix.rda")
ef <- ef_fix
rm(ef_fix)
GMid <- unique(ef$GrandmotherID)
Mid <- unique(ef$MotherID)
Chid <- unique(ef$Id_I) 
load("../IDS/IDS_ddb/INDIVIDUAL.rda") 
INDIVIDUAL <- INDIVIDUAL %>%
   select(Id_I, Type, Value, Day, Month, Year, Start_day, Start_month, 
           Start_year, End_day, End_month, End_year,
           Date_type, Missing)
load("../IDS/IDS_ddb/INDIV_INDIV.rda") 
INDIV_INDIV <- INDIV_INDIV %>%
    select(Id_I_1, Id_I_2, Relation) %>%
    filter(Relation == "Mother") %>% 
    select(Mother = Id_I_1, Child = Id_I_2) %>%
    filter(!duplicated(Child))
```

Now, from *INDIV_INDIV* we select all rows ('filter') with *Mother* 
corresponding to *GMid* (from *ef*).

```{r Gmchild}
GMchild <- INDIV_INDIV %>%
    filter(Mother %in% GMid)
```

These grandmothers had a lot of children:

```{r tapply}
x <- with(GMchild, tapply(Mother, Mother, length))
barplot(table(x), xlab = "No. of children to grandmother", ylab = "Frequency")
```

## Find the kids of the grandmothers

That is, the *mothers* and their siblings. From *INDIVIDUAL*, after some 
massage:

```{r get_kids_to_GM}
kidsGM <- INDIVIDUAL %>%
    filter(Id_I %in% GMchild$Child) %>% 
    filter(Type %in% c("BIRTH_DATE", "DEATH_DATE", "START_OBSERVATION", 
           "END_OBSERVATION"))
kidsGM$date <- with(kidsGM, mdy(Month, Day, Year))
kidsGM <- select(kidsGM, Id_I, Type, Value, date) %>%
    filter(!is.na(date))

far <- with(kidsGM, paste(Id_I, Type, Value, date, sep = ""))
kidsGM <- kidsGM[!duplicated(far), ]
                
indx <- match(kidsGM$Id_I, INDIV_INDIV$Child)
kidsGM$GMid <- INDIV_INDIV$Mother[indx]
summary(kidsGM)
```

We see that grandmother has a birth as early as 1804-01-13 and her longest 
living kid dies 1902-01-026.

We will now check the birthdate , deathdate and 'last seen' for all GMkids.
We start by sorting data by Id_I and date and then add *lopnr* and *antrec*: 

```{r rc}
source("rc.R")
kidsGM$id <- kidsGM$Id_I
kidsGM <- rc(kidsGM)
kidsGM$id <- NULL
head(kidsGM, 15)
```

Individuals with missing birthdate are removed:

```{r withbdate}
id <- kidsGM$Id_I[kidsGM$Type == "BIRTH_DATE"]
kids <- kidsGM %>% 
    filter(Id_I %in% id)
kids$id <- kids$Id_I
kids <- rc(kids)
kids$id <- NULL
noval <- kids$Value == ""
kids$Value[noval] <- as.character(kids$date[noval])
kk <- spread(kids, Type, Value)
kk$
head(kk)
```



